---
title: "Analyzing Word Rating Studies with Linear Mixed-Effects Models (lmer)"
author: "Introduction to R for Psycholinguistics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

# Introduction

Many experiments in psycholinguistics ask participants to rate words or
sentences using a continuous scale.

Examples include ratings of

- valence
- arousal
- dominance
- concreteness
- familiarity
- imageability
- semantic similarity
- acceptability

A common method is a **100-point slider**, where participants move a slider
between two labeled endpoints.

Because every participant rates many words and every word is rated by many
participants, the observations are **not independent**.

Linear mixed-effects models (LMMs) provide an appropriate statistical method
for analyzing these data.

In this tutorial you will learn how to

- import rating data
- inspect the data
- visualize distributions
- fit mixed-effects models using `lmer()`
- interpret model output
- compare competing models
- obtain estimated marginal means
- visualize model predictions

---

# Packages

```{r message=FALSE}

library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)

```

---

# Example Dataset

Suppose participants rated words for emotional valence.

Each participant rated many words.

Our dataset contains

| Variable | Description |
|-----------|-------------|
| Participant | Participant ID |
| Word | Target word |
| Rating | Slider response (0–100) |
| Condition | Experimental manipulation |
| Valence | Lexical valence |
| Frequency | Word frequency |

---

# Loading the Data

```{r}

ratings <- read_csv("word_ratings.csv")

glimpse(ratings)

```

---

# Inspecting the Data

```{r}

summary(ratings)

```

Look for

- missing values
- unexpected ranges
- coding errors

---

# Distribution of Ratings

```{r}

ggplot(ratings,
       aes(Rating)) +

geom_histogram(bins = 30) +

theme_minimal()

```

Questions:

- Are ratings approximately normal?
- Are there ceiling or floor effects?

---

# Ratings by Condition

```{r}

ggplot(ratings,

aes(Condition,
    Rating)

) +

geom_boxplot()

```

---

# Individual Participant Means

```{r}

ratings %>%

group_by(Participant) %>%

summarise(

MeanRating = mean(Rating)

)

```

---

# Why Mixed Models?

Traditional ANOVAs assume observations are independent.

In word-rating experiments,

- participants differ in their rating behavior
- words differ in their inherent properties

Mixed models account for both sources of variability.

---

# Random Effects

Our observations are clustered by

- Participant
- Word

Therefore we include random intercepts for both.

Mathematically,

```
Rating ~ Predictors +
         (1 | Participant) +
         (1 | Word)
```

---

# Fitting the Simplest Model

```{r}

m0 <-

lmer(

Rating ~

1 +

(1 | Participant) +

(1 | Word),

data = ratings

)

summary(m0)

```

This model estimates only the grand mean.

---

# Adding an Experimental Condition

Suppose words belong to two conditions.

```{r}

m1 <-

lmer(

Rating ~

Condition +

(1 | Participant) +

(1 | Word),

data = ratings

)

summary(m1)

```

---

# Interpreting Fixed Effects

The output includes

- Estimate
- Standard Error
- t value
- p value (provided by **lmerTest**)

For example,

```
ConditionB = 4.25
```

means ratings in Condition B are estimated to be 4.25 points higher than the
reference condition.

---

# Adding Continuous Predictors

Suppose lexical valence also predicts ratings.

```{r}

m2 <-

lmer(

Rating ~

Condition +

Valence +

(1 | Participant) +

(1 | Word),

data = ratings

)

summary(m2)

```

---

# Interactions

Perhaps lexical valence matters differently across conditions.

```{r}

m3 <-

lmer(

Rating ~

Condition * Valence +

(1 | Participant) +

(1 | Word),

data = ratings

)

summary(m3)

```

---

# Comparing Models

Does the interaction improve the model?

```{r}

anova(

m2,
m3

)

```

Smaller AIC values and significant likelihood-ratio tests indicate improved fit.

---

# Random Slopes

Participants may respond differently to the experimental manipulation.

```{r}

m4 <-

lmer(

Rating ~

Condition +

(1 + Condition | Participant) +

(1 | Word),

data = ratings

)

summary(m4)

```

Random slopes allow the effect of Condition to vary across participants.

---

# Model Diagnostics

Check residuals.

```{r}

plot(m3)

```

Look for

- equal variance
- no systematic patterns

---

# QQ Plot

```{r}

qqnorm(residuals(m3))

qqline(residuals(m3))

```

Residuals should approximately follow a straight line.

---

# Estimated Marginal Means

Estimated marginal means summarize the predicted means after accounting for
other variables in the model.

```{r}

emmeans(

m3,

~ Condition

)

```

---

# Pairwise Comparisons

```{r}

pairs(

emmeans(

m3,

~ Condition

)

)

```

---

# Plotting Model Predictions

```{r}

emm <-

emmeans(

m3,

~ Condition

)

plot(emm)

```

---

# Visualizing Raw Data

```{r}

ggplot(

ratings,

aes(

Condition,

Rating

)

) +

geom_violin(fill = "lightblue") +

geom_boxplot(width = .15)

```

---

# Participant-Level Means

```{r}

participant_means <-

ratings %>%

group_by(

Participant,

Condition

) %>%

summarise(

Mean = mean(Rating),

.groups = "drop"

)

ggplot(

participant_means,

aes(

Condition,

Mean

)

) +

geom_point(alpha = .4)

```

---

# Adding Multiple Predictors

A more realistic psycholinguistic model might include

```{r}

full_model <-

lmer(

Rating ~

Condition +

Valence +

Frequency +

Length +

Condition:Valence +

(1 | Participant) +

(1 | Word),

data = ratings

)

summary(full_model)

```

---

# Interpreting Coefficients

Suppose we obtain

| Predictor | Estimate |
|------------|----------|
| Condition | 5.2 |
| Valence | 3.4 |
| Frequency | -1.1 |

Interpretation:

- Condition increases ratings by approximately five points.
- Higher-valence words receive higher ratings.
- More frequent words receive slightly lower ratings.

---

# Reporting Results

A typical report might read:

> A linear mixed-effects model was fit using the **lme4** package (Bates et al.,
2015). Fixed effects included Condition and lexical Valence. Random intercepts
were included for Participant and Word. Condition significantly affected ratings
(β = 5.21, SE = 1.03, *t* = 5.07, *p* < .001).

---

# Common Random Effects Structures

| Formula | Interpretation |
|----------|----------------|
| `(1 | Participant)` | Random participant intercepts |
| `(1 | Word)` | Random word intercepts |
| `(1 + Condition | Participant)` | Participant-specific condition effects |
| `(1 + Valence | Participant)` | Participant-specific valence effects |
| `(1 + Condition | Word)` | Word-specific condition effects |

---

# Common Problems

## Convergence Warnings

Possible solutions:

- simplify the random-effects structure
- center continuous predictors
- increase optimizer iterations

```{r}

ratings <-

ratings %>%

mutate(

Valence_c = scale(Valence,
                   scale = FALSE)

)

```

---

## Singular Fits

A singular fit often indicates that the model is attempting to estimate more
random-effect parameters than the data can support.

Simplify the random-effects structure if necessary.

---

# Exercises

## Exercise 1

Fit a model predicting ratings from Condition.

---

## Exercise 2

Add lexical Valence.

Does model fit improve?

---

## Exercise 3

Fit an interaction between Condition and Valence.

Interpret the interaction.

---

## Exercise 4

Add random slopes for participants.

Does the model converge?

---

## Exercise 5

Use `emmeans()` to compare conditions.

---

## Exercise 6

Create a figure showing the estimated marginal means with confidence intervals.

---

# Summary

In this tutorial you learned how to

- inspect word-rating data
- visualize rating distributions
- fit mixed-effects models
- compare nested models
- interpret fixed effects
- specify random intercepts and random slopes
- diagnose model assumptions
- estimate marginal means
- visualize model predictions

Linear mixed-effects models have become the standard approach for analyzing
continuous psycholinguistic rating data because they account for variability
across both participants and items while allowing multiple experimental and
lexical predictors to be modeled simultaneously.