---
title: "Corpus Analysis with tidytext"
author: "Introduction to R for Data Science"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

# Introduction

In the previous tutorial we learned how to:

- import a corpus
- tokenize text
- count word frequencies
- merge lexical norms
- examine emotional properties using the Warriner et al. norms

In this lesson we'll explore some of the most common techniques used in
corpus linguistics and text mining with the **tidytext** package.

By the end of this tutorial you will be able to:

- remove stopwords
- examine word frequencies
- create bar charts
- calculate TF-IDF
- identify common word pairs (bigrams)
- compare multiple documents
- perform sentiment analysis using several popular lexicons

---

# Packages

```{r message=FALSE}

library(tidyverse)
library(tidytext)

```

---

# Loading a corpus

```{r}

documents <- tibble(

doc = c(

"Article 1",
"Article 2",
"Article 3"

),

text = c(

"The puppy happily chased the ball through the garden.",

"The politician announced difficult economic reforms.",

"The students celebrated after finishing their final exams."

)

)

documents

```

---

# Tokenization

```{r}

tokens <-

documents %>%

unnest_tokens(word, text)

tokens

```

Each row now represents a single word.

---

# Word frequencies

```{r}

word_counts <-

tokens %>%

count(word, sort = TRUE)

word_counts

```

---

# Plotting the most common words

```{r}

word_counts %>%

slice_max(n, n = 10) %>%

ggplot(aes(reorder(word, n), n)) +

geom_col() +

coord_flip() +

labs(

x = "",

y = "Frequency",

title = "Most Common Words"

)

```

---

# Stopwords

Many common words (such as *the*, *and*, *of*) occur frequently but carry
little semantic information.

The tidytext package includes a built-in stopword list.

```{r}

stop_words

```

---

# Removing stopwords

```{r}

tokens_clean <-

tokens %>%

anti_join(stop_words,
          by = "word")

tokens_clean

```

---

# Word frequencies after cleaning

```{r}

tokens_clean %>%

count(word, sort = TRUE)

```

Notice how the informative words become much easier to identify.

---

# Comparing documents

```{r}

tokens_clean %>%

count(doc, word)

```

Now we know how often every word appears in every document.

---

# TF-IDF

Raw frequency is useful, but common words may dominate every document.

TF-IDF (Term Frequency–Inverse Document Frequency) identifies words that are
particularly characteristic of individual documents.

```{r}

tfidf <-

tokens_clean %>%

count(doc, word) %>%

bind_tf_idf(word, doc, n)

tfidf

```

---

# Highest TF-IDF words

```{r}

tfidf %>%

group_by(doc) %>%

slice_max(tf_idf, n = 5)

```

These are the words that best distinguish each document.

---

# Visualizing TF-IDF

```{r}

tfidf %>%

group_by(doc) %>%

slice_max(tf_idf, n = 5) %>%

ungroup() %>%

ggplot(aes(reorder_within(word,
                          tf_idf,
                          doc),
           tf_idf)) +

geom_col() +

facet_wrap(~doc,
           scales = "free") +

scale_x_reordered() +

coord_flip()

```

---

# Sentiment analysis

The tidytext package includes several sentiment lexicons.

Let's examine them.

```{r}

get_sentiments("bing")

get_sentiments("afinn")

get_sentiments("nrc")

```

---

# Bing lexicon

The Bing lexicon labels words as either

- positive
- negative

```{r}

bing_words <-

tokens_clean %>%

inner_join(get_sentiments("bing"),
           by = "word")

bing_words

```

---

# Counting positive and negative words

```{r}

bing_words %>%

count(sentiment)

```

---

# AFINN

The AFINN lexicon assigns each word a score from

-5 (very negative)

to

+5 (very positive)

```{r}

afinn <-

tokens_clean %>%

inner_join(get_sentiments("afinn"),
           by = "word")

afinn

```

---

# Average sentiment

```{r}

afinn %>%

summarise(

Mean_Sentiment = mean(value)

)

```

---

# NRC Emotion Lexicon

The NRC lexicon classifies words into emotions such as

- joy
- anger
- fear
- trust
- sadness
- surprise
- disgust
- anticipation

```{r}

nrc <-

tokens_clean %>%

inner_join(get_sentiments("nrc"),
           by = "word")

head(nrc)

```

---

# Emotion counts

```{r}

nrc %>%

count(sentiment)

```

---

# Plotting emotions

```{r}

nrc %>%

count(sentiment) %>%

ggplot(aes(reorder(sentiment, n), n)) +

geom_col() +

coord_flip()

```

---

# Bigrams

Instead of looking at individual words, we can examine pairs of words.

```{r}

bigrams <-

documents %>%

unnest_tokens(bigram,
              text,
              token = "ngrams",
              n = 2)

bigrams

```

Examples include

- economic reforms
- final exams

---

# Separating bigrams

```{r}

bigrams_sep <-

bigrams %>%

separate(

bigram,

into = c("word1", "word2"),

sep = " "

)

bigrams_sep

```

---

# Removing stopwords from bigrams

```{r}

bigrams_clean <-

bigrams_sep %>%

filter(

!word1 %in% stop_words$word,

!word2 %in% stop_words$word

)

bigrams_clean

```

---

# Counting bigrams

```{r}

bigrams_clean %>%

count(word1,
      word2,
      sort = TRUE)

```

---

# Comparing sentiment methods

| Lexicon | Output |
|----------|---------|
| Bing | Positive / Negative |
| AFINN | Numeric score |
| NRC | Eight emotions + positive/negative |
| Warriner | Continuous Valence, Arousal, Dominance |

Each captures a different aspect of emotional language.

---

# Putting everything together

A typical corpus analysis workflow looks like this:

```
Import text

↓

Tokenize

↓

Remove stopwords

↓

Count frequencies

↓

Calculate TF-IDF

↓

Join lexical norms

↓

Calculate sentiment

↓

Visualize results

↓

Interpret findings
```

---

# Exercises

## Exercise 1

Remove stopwords from your own corpus.

How do the most frequent words change?

---

## Exercise 2

Calculate TF-IDF for your corpus.

Which words are most characteristic of each document?

---

## Exercise 3

Compare the Bing and AFINN lexicons.

Do they produce similar conclusions?

---

## Exercise 4

Create a bigram table.

Which word pairs occur most often?

---

## Exercise 5

Join the Warriner norms from the previous tutorial.

Compare the average

- Valence
- Arousal
- Dominance

for each document.

---

# Summary

You have now learned how to

- tokenize text
- remove stopwords
- count word frequencies
- calculate TF-IDF
- use three common sentiment lexicons
- extract bigrams
- compare multiple documents
- combine lexical resources with corpus analysis

These techniques form the foundation of modern corpus linguistics,
computational social science, digital humanities, and natural language
processing.