---
title: "Corpus Analysis with Warriner et al. Norms"
author: "morgan moyer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---



1.  Affect Norms:
>[Warriner, Kuperman & Brysbaert (2013)](https://link.springer.com/article/10.3758/s13428-012-0314-x)

2. Concreteness Norms:
> [Brysbaert, Warriner, Kuperman (2014)](https://link.springer.com/article/10.3758/s13428-013-0403-5)

3. Word Frequency corpus:
> [Subtlex Corpus](https://www.ugent.be/pp/experimentele-psychologie/en/research/documents/subtlexus)

# Packages

```{r message=FALSE}

library(tidyverse)
library(tidytext)
library(ggplot2)


# Set working directory to the current directory
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)
```


# Loading the corporas

```{r}

war <- read.csv("../data/warriner.csv")
brys <- read.csv("../data/brysbaert.csv")
subtlex <- read.csv("../data/SUBTLEX-US.csv")

head(war)
head(subtlex)
```


The important columns are:

```{r}

names(war)
names(brys)
names(subtlex)

```


Rename the columns to something easier to use.

```{r}

war <- war %>%
      transmute(
        Word = Word,
        valence = V.Mean.Sum,
        arousal = A.Mean.Sum,
        dominance = D.Mean.Sum,
        valence.sd = V.SD.Sum,
        arousal.sd = A.SD.Sum,
        dominance.sd = D.SD.Sum
      
      )

head(war)

```


# Distribution of valence

```{r}

ggplot(war,
       aes(valence)) +

geom_histogram(bins = 30)

```

Questions:

- Is the corpus generally positive?
- Are there many emotionally neutral words?

---

# Distribution of arousal

```{r}

ggplot(war,
       aes(arousal)) +

geom_histogram(bins = 30)

```

---

# Distribution of dominance

```{r}

ggplot(war,
       aes(dominance)) +

geom_histogram(bins = 30)

```


# take a look at valence and arousal as a 

```{r}
ggplot(war, aes(x = arousal, y = valence)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Arousal (A.Mean.Sum)",
    y = "Valence (V.Mean.Sum)",
    title = "Valence as a Function of Arousal"
  )
```

## calculate correlation
```{r}
cor(war$arousal, war$valence, use = "complete.obs")
```


# Add word labels
```{r}

head(war)
ggplot(war, aes(x = arousal, y = valence, label = Word)) +
  geom_text(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Arousal (A.Mean.Sum)",
    y = "Valence (V.Mean.Sum)",
    title = "Valence as a Function of Arousal"
  )
```


# Frequency
```{r}
# subset data just to make it more managable
subtlex <- subtlex[,c("Word","SUBTLWF")]
head(subtlex)
```


# Joining norms together

Merge the datasets by Word
```{r}

total <- war |>  
  left_join(brys,
          by = "Word") |> 
  left_join(subtlex, by="Word")

head(total)

```

Many words will have ratings.

Some will not.

---

# Missing values

How many words have ratings?

```{r}

sum(!is.na(total$Conc.M))

sum(is.na(total$Conc.M))

length(unique(total$Word))

nrow(total)
```

## Drop all words that don't have ratings for both concreteness and affect
```{r}
total <- total |> 
  filter(!is.na(Conc.M))
nrow(total)
```

# Average emotional properties

Now calculate the average emotional profile of the corpus.

```{r}

agr <- total %>%
      summarise(
          Mean_Valence = mean(valence, na.rm = TRUE),
          Mean_Arousal = mean(arousal, na.rm = TRUE),
          Mean_Dominance = mean(dominance, na.rm = TRUE),
          Mean_Conc = mean(Conc.M, na.rm = TRUE)
      )

print(agr)
```

# Which emotional words are most frequent?

```{r}
names(total)
total %>%
  arrange(desc(SUBTLWF)) %>%
  head()

```

```{r}
total %>%
  arrange(desc(SUBTLWF)) %>%
  tail()
```

---

# Most positive words

```{r}

total %>%
  arrange(desc(valence)) |> 
  head()

```

---

# Most negative words

```{r}

total %>%
  arrange(desc(valence)) |> 
  tail()

```

---

# Relationship between Valence and Arousal

```{r}

ggplot(total,
       aes(valence,
           arousal)) +

geom_point(alpha = .3)

```

Questions:

- Are positive words highly arousing?
- Are negative words highly arousing?



# Concreteness

```{r}
ggplot(total,
       aes(valence,
           Conc.M)) +

geom_point(alpha = .3)

```

```{r}
ggplot(total, aes(x = Conc.M, y = valence)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  # theme_minimal() +
  labs(
    x = "Concreteness",
    y = "Valence (V.Mean.Sum)",
    title = "Valence as a Function of Concreteness"
  )
```

# Frequency and Valence?

```{r}
ggplot(total, aes(x = SUBTLWF, y = valence, label=Word)) +
  geom_text(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  # theme_minimal() +
  labs(
    x = "Frequency",
    y = "Valence",
    title = "Valence as a Function of Frequency"
  )
```


# Frequency and Arousal?

```{r}
ggplot(total, aes(x = SUBTLWF, y = arousal, label=Word)) +
  geom_text(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  # theme_minimal() +
  labs(
    x = "Frequency",
    y = "Arousal",
    title = "Arousal as a Function of Frequency"
  )
```

# Exercises

### Exercise 1

Calculate the average variance in valence.

---

### Exercise 2

Which ten words contribute most to the emotional profile?

---

### Exercise 3

Compare two words

Which is:

- more positive?
- more arousing?
- more dominant?

---

### Exercise 4

Create a scatterplot of valence versus dominance.

What relationship do you observe?

---