---
title: "Corpus Analysis with Warriner et al. Norms"
author: "morgan moyer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

# Introduction

One of the strengths of computational linguistics is the ability to combine
large text corpora with lexical databases.

In this tutorial we will learn how to:

- Import a corpus
- Tokenize text into words
- Calculate word frequencies
- Examine the  Warriner et al. (2013) norms of **Valence**, **Arousal**, and **Dominance**
- Visualize
---

# What are lexical norms?

Lexical norms are databases in which human participants have rated words
along different psychological dimensions.

One of the most widely used datasets is:

> Warriner, Kuperman & Brysbaert (2013)

Participants rated nearly 14,000 English words on three scales.

| Dimension | Meaning |
|------------|----------|
| **Valence** | How pleasant or unpleasant a word is |
| **Arousal** | How exciting or calming a word is |
| **Dominance** | How much control or power is associated with a word |

Examples:

| Word | Valence | Arousal | Dominance |
|------|---------|----------|------------|
| love | High | Medium | Medium |
| death | Low | High | Low |
| puppy | High | Medium | Low |
| dictator | Low | High | High |

---

# Packages

```{r message=FALSE}

library(tidyverse)
library(tidytext)

# Set working directory to the current directory
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)
```

---

# Loading the corpora

For this example we'll use a simple text file.

```{r}

war <- read.csv("../data/warriner.csv")
brys <- read.csv("../data/brysbaert.csv")
subtlex <- read.csv("../data/SUBTLEX-US.csv")

head(war)

```


The important columns are:

```{r}

names(war)

```

Typically the variables include

- Word
- V.Mean.Sum
- A.Mean.Sum
- D.Mean.Sum

---

# Preparing the norms

Rename the columns to something easier to use.

```{r}

war<-

war%>%

transmute(

word = Word,

valence = V.Mean.Sum,

arousal = A.Mean.Sum,

dominance = D.Mean.Sum

)

head(war)

```

---

# Joining lexical norms to the corpus

Now we merge the emotional ratings onto each token.

```{r}

tokens_va <-

tokens %>%

left_join(war,
          by = "word")

head(tokens_va)

```

Many words will have ratings.

Some will not.

---

# Missing values

How many words have ratings?

```{r}

mean(!is.na(tokens_va$valence))

```

This gives the proportion of corpus words covered by the norms.

---

# Average emotional properties

Now calculate the average emotional profile of the corpus.

```{r}

tokens_va %>%

summarise(

Mean_Valence = mean(valence, na.rm = TRUE),

Mean_Arousal = mean(arousal, na.rm = TRUE),

Mean_Dominance = mean(dominance, na.rm = TRUE)

)

```

---

# Distribution of valence

```{r}

ggplot(tokens_va,
       aes(valence)) +

geom_histogram(bins = 30)

```

Questions:

- Is the corpus generally positive?
- Are there many emotionally neutral words?

---

# Distribution of arousal

```{r}

ggplot(tokens_va,
       aes(arousal)) +

geom_histogram(bins = 30)

```

---

# Distribution of dominance

```{r}

ggplot(tokens_va,
       aes(dominance)) +

geom_histogram(bins = 30)

```

---

# Which emotional words are most frequent?

```{r}

tokens_va %>%

filter(!is.na(valence)) %>%

count(word,
      valence,
      sort = TRUE)

```

---

# Most positive words

```{r}

tokens_va %>%

distinct(word, valence) %>%

arrange(desc(valence))

```

---

# Most negative words

```{r}

tokens_va %>%

distinct(word, valence) %>%

arrange(valence)

```

---

# Relationship between Valence and Arousal

```{r}

ggplot(tokens_va,
       aes(valence,
           arousal)) +

geom_point(alpha = .3)

```

Questions:

- Are positive words highly arousing?
- Are negative words highly arousing?

---

# Looking at individual texts

Suppose each row of the corpus is a document.

```{r}

documents <-

tibble(

doc = 1:length(corpus),

text = corpus

)

```

Tokenize while keeping the document ID.

```{r}

doc_tokens <-

documents %>%

unnest_tokens(word, text) %>%

left_join(war,
          by = "word")

```

Now compute emotional averages for each document.

```{r}

document_scores <-

doc_tokens %>%

group_by(doc) %>%

summarise(

valence = mean(valence, na.rm = TRUE),

arousal = mean(arousal, na.rm = TRUE),

dominance = mean(dominance, na.rm = TRUE)

)

document_scores

```

---

# Visualizing documents

```{r}

ggplot(document_scores,
       aes(valence,
           arousal)) +

geom_point(size = 3)

```

Each point represents one document.

---

# Discussion

Lexical norms allow us to estimate the emotional characteristics of texts
without asking participants to rate every document.

This approach is widely used in:

- sentiment analysis
- computational linguistics
- digital humanities
- psychology
- communication science

However, remember that lexical norms have limitations.

For example:

- Words can have different meanings in different contexts.
- Negation ("not happy") changes emotional meaning.
- Irony and sarcasm are difficult to detect.
- Emotional meaning is not simply the average of individual words.

---

# Exercises

### Exercise 1

Calculate the average valence of your own corpus.

---

### Exercise 2

Which ten words contribute most to the emotional profile?

---

### Exercise 3

Compare two texts.

Which is:

- more positive?
- more arousing?
- more dominant?

---

### Exercise 4

Create a scatterplot of valence versus dominance.

What relationship do you observe?

---

# Summary

In this tutorial you learned how to:

- tokenize text
- count word frequencies
- merge lexical norms with corpus data
- compute emotional summaries
- visualize emotional properties of texts
- compare documents using lexical ratings

These techniques form the basis of many modern approaches to sentiment
analysis and computational psycholinguistics.