---
title: "Basics of Probability Theory for Statistical Analysis"
author: "Introduction to Statistics for Psycholinguistics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

# Introduction

Probability theory provides the mathematical foundation for statistical
analysis.

Whenever we conduct an experiment, we are dealing with uncertainty.

For example:

- Will a participant give a high or low word rating?
- Will a participant respond correctly?
- Will a particular word occur in a corpus?
- Will an experimental manipulation produce a difference?
- How much would our results vary if we repeated the experiment?

Statistics gives us tools for learning about populations from samples.

Probability theory gives us a language for describing uncertainty.

In this tutorial we will develop the basic ideas of probability that underlie
statistical analysis.

By the end of this tutorial you should understand:

- random experiments
- outcomes and events
- probability
- conditional probability
- independence
- random variables
- probability distributions
- expectation and variance
- the binomial distribution
- the normal distribution
- sampling and repeated experiments
- the connection between probability and statistical inference

We will use R to simulate many of these concepts.

---

# Packages

```{r message=FALSE}

library(tidyverse)

```

---

# 1. Random Experiments

A **random experiment** is an experiment whose outcome cannot be predicted
with certainty in advance.

Examples include:

- rolling a die
- flipping a coin
- selecting a word from a corpus
- selecting a participant
- asking a participant to rate a word

The important point is that we may know the possible outcomes without knowing
which outcome will occur.

---

# 2. Outcomes and Sample Spaces

An **outcome** is a possible result of an experiment.

For example, when rolling a six-sided die:

```text
1
2
3
4
5
6
```

The **sample space** is the set of all possible outcomes.

We can represent it in R:

```{r}

die <- 1:6

die

```

---

# 3. Events

An **event** is a set of possible outcomes.

For example:

> The die shows an even number.

The event is:

```{r}

even <- c(2, 4, 6)

even

```

Another event is:

> The die shows a number greater than 4.

```{r}

greater_than_4 <- c(5, 6)

greater_than_4

```

An event can contain one outcome or many outcomes.

---

# 4. Probability

Probability describes how likely an event is.

The probability of an event is always between 0 and 1.

```text
0 = impossible

1 = certain
```

For a fair six-sided die:

\[
P(1) = \frac{1}{6}
\]

and

\[
P(\text{even}) = \frac{3}{6} = 0.5
\]

---

# 5. Calculating Probabilities

If all outcomes are equally likely:

\[
P(A) =
\frac{\text{number of outcomes in } A}
{\text{number of possible outcomes}}
\]

For an even number:

```{r}

length(even) / length(die)

```

For a number greater than 4:

```{r}

length(greater_than_4) / length(die)

```

---

# 6. Simulating Probability

Instead of calculating probabilities analytically, we can simulate experiments.

Roll a die once:

```{r}

sample(die,
       size = 1)

```

Roll it 10 times:

```{r}

sample(die,
       size = 10,
       replace = TRUE)

```

The argument `replace = TRUE` means that the same outcome can occur more
than once.

---

# 7. The Law of Large Numbers

Let's roll the die 10 times.

```{r}

rolls_10 <-

sample(

die,

size = 10,

replace = TRUE

)

mean(rolls_10)

```

Now try 10,000 rolls.

```{r}

rolls_10000 <-

sample(

die,

size = 10000,

replace = TRUE

)

mean(rolls_10000)

```

The average should approach the theoretical expected value of a fair die:

\[
E(X) = 3.5
\]

This illustrates the **Law of Large Numbers**.

As the number of observations increases, the sample average tends to approach
the population expectation.

---

# 8. Simulating a Probability

Suppose we want to estimate:

\[
P(\text{even})
\]

Simulate 10,000 rolls.

```{r}

rolls <-

sample(

die,

size = 10000,

replace = TRUE

)

mean(rolls %% 2 == 0)

```

The result should be close to 0.5.

We have estimated a probability through simulation.

---

# 9. Repeated Experiments

Let's repeat the simulation.

```{r}

simulation <-

replicate(

100,

mean(

sample(

die,

size = 100,

replace = TRUE

) %% 2 == 0

)

)

simulation

```

Look at the distribution.

```{r}

tibble(
  proportion_even = simulation
) %>%

ggplot(aes(proportion_even)) +

geom_histogram(bins = 15) +

theme_minimal()

```

Notice that the estimated probability varies from experiment to experiment.

This idea will become extremely important later.

---

# 10. Probability and Uncertainty

Suppose the true probability is:

\[
p = 0.5
\]

We do not expect every sample to contain exactly 50% even rolls.

Instead, different samples produce different estimates.

This is **sampling variability**.

Statistical inference is largely concerned with understanding this variability.

---

# 11. Complementary Events

The **complement** of an event is everything that is not in that event.

If:

\[
P(A) = 0.7
\]

then:

\[
P(A^c) = 1 - P(A)
\]

For example:

\[
P(\text{even}) = 0.5
\]

Therefore:

\[
P(\text{not even}) = 0.5
\]

---

# 12. Union of Events

The union of two events means:

> A or B occurs.

We write:

\[
P(A \cup B)
\]

For example:

- A = die is even
- B = die is greater than 4

The relevant outcomes are:

```text
2, 4, 5, 6
```

Therefore:

\[
P(A \cup B) = \frac{4}{6}
\]

---

# 13. Intersection of Events

The intersection means:

> A and B both occur.

We write:

\[
P(A \cap B)
\]

For the die:

- A = even
- B = greater than 4

Only 6 satisfies both.

Therefore:

\[
P(A \cap B) = \frac{1}{6}
\]

---

# 14. Conditional Probability

Conditional probability is one of the most important concepts in statistics.

It asks:

> What is the probability of A given that we know B has occurred?

We write:

\[
P(A \mid B)
\]

This is read:

> "The probability of A given B."

---

# 15. An Example

Suppose we roll a die.

What is the probability of getting an even number?

\[
P(\text{even}) = \frac{3}{6}
\]

But suppose we know that the die is greater than 3.

The possible outcomes are now:

```text
4, 5, 6
```

Of these, two are even.

Therefore:

\[
P(\text{even} \mid >3)
=
\frac{2}{3}
\]

The information we have changes the probability.

---

# 16. The Conditional Probability Formula

The general formula is:

\[
P(A \mid B)
=
\frac{P(A \cap B)}
{P(B)}
\]

This formula is fundamental to statistical reasoning.

---

# 17. Conditional Probability in Research

Conditional probability appears everywhere in empirical research.

For example:

> What is the probability that a participant gives a high rating given that
> the word has high valence?

Or:

> What is the probability that a word occurs given that it belongs to a
> particular genre?

Or:

> What is the probability of observing our data given a particular statistical
> model?

The final example leads directly toward hypothesis testing.

---

# 18. Independence

Two events are **independent** if knowing that one occurred does not change
the probability of the other.

For independent events:

\[
P(A \mid B) = P(A)
\]

Equivalently:

\[
P(A \cap B) = P(A)P(B)
\]

---

# 19. Example: Coin Flips

Suppose we flip a fair coin twice.

The first flip does not affect the second.

Therefore:

\[
P(\text{Heads on flip 2} \mid
\text{Heads on flip 1})
=
P(\text{Heads on flip 2})
=
0.5
\]

The flips are independent.

---

# 20. Independence in Experimental Data

Independence is an important assumption in statistics.

However, observations in psycholinguistic experiments are often **not**
independent.

For example, if one participant rates 100 words:

```text
Participant 1
  ├── Word 1
  ├── Word 2
  ├── Word 3
  ...
  └── Word 100
```

Those observations come from the same participant.

Similarly, the same word may be rated by many participants.

This is one reason mixed-effects models are useful.

---

# 21. Random Variables

A **random variable** is a variable whose value depends on the outcome of a
random experiment.

For example:

\[
X = \text{result of a die roll}
\]

Possible values are:

```text
1, 2, 3, 4, 5, 6
```

Another example:

\[
Y = \text{rating given by a participant}
\]

Here the possible values might range from:

```text
0 to 100
```

---

# 22. Discrete and Continuous Variables

Random variables can be broadly divided into two types.

### Discrete

A discrete variable takes separate, countable values.

Examples:

- number of correct responses
- number of words
- number of errors
- die roll

### Continuous

A continuous variable can take values on a continuous scale.

Examples:

- reaction time
- slider rating
- response time
- height

---

# 23. Probability Distributions

A probability distribution describes how probability is distributed across the
possible values of a random variable.

For a fair die:

```text
Value     Probability

1         1/6
2         1/6
3         1/6
4         1/6
5         1/6
6         1/6
```

---

# 24. Expected Value

The **expected value** is the long-run average value of a random variable.

For a discrete variable:

\[
E(X) = \sum_x xP(X=x)
\]

For a fair die:

\[
E(X)
=
1(1/6)+2(1/6)+...+6(1/6)
\]

which equals:

\[
3.5
\]

Importantly, 3.5 is not actually an outcome of the die.

It is the long-run average.

---

# 25. Variance

Expected value describes the center of a distribution.

Variance describes how spread out the distribution is.

\[
Var(X) =
E[(X-E(X))^2]
\]

The standard deviation is:

\[
SD(X) = \sqrt{Var(X)}
\]

---

# 26. Why Variance Matters

Imagine two populations.

Population A:

```text
48 49 50 51 52
```

Population B:

```text
10 30 50 70 90
```

Both have approximately the same mean.

But Population B has much greater variability.

Statistical inference must account for both:

- the mean
- the variability

---

# 27. The Binomial Distribution

The binomial distribution describes the number of successes in a fixed number
of independent trials when each trial has the same probability of success.

For example:

> A participant has a 70% probability of answering correctly. What is the
> probability of getting exactly 7 correct responses out of 10?

---

# 28. Simulating Binomial Data

```{r}

rbinom(

n = 20,

size = 10,

prob = 0.7

)

```

Each number represents the number of successes in 10 trials.

---

# 29. Visualizing the Binomial Distribution

```{r}

binomial_data <-

tibble(

successes = rbinom(

10000,

size = 10,

prob = 0.7

)

)

ggplot(

binomial_data,

aes(successes)

) +

geom_bar() +

theme_minimal()

```

---

# 30. The Normal Distribution

The normal distribution is one of the most important distributions in
statistics.

It is characterized by:

- a mean
- a standard deviation

```{r}

x <-

seq(-4, 4,
    length.out = 500)

normal_data <-

tibble(

x = x,

density = dnorm(x)

)

ggplot(

normal_data,

aes(x,
    density)

) +

geom_line() +

theme_minimal()

```

---

# 31. Standard Normal Distribution

A standard normal distribution has:

\[
\mu = 0
\]

and

\[
\sigma = 1
\]

We can generate values from it using:

```{r}

rnorm(

1000,

mean = 0,

sd = 1

)

```

---

# 32. Z-Scores

A z-score tells us how many standard deviations an observation is from the
mean.

\[
z =
\frac{x-\mu}{\sigma}
\]

For example, if:

\[
\mu = 50
\]

and:

\[
\sigma = 10
\]

then a score of 70 has:

\[
z = \frac{70-50}{10} = 2
\]

The observation is two standard deviations above the mean.

---

# 33. Sampling

Suppose there is a population of 100,000 possible observations.

We usually cannot measure everyone.

Instead, we collect a sample.

```text
Population
    |
    | sample
    v
Sample
    |
    | analysis
    v
Inference about population
```

This is the fundamental logic of statistical inference.

---

# 34. Sampling Variability

Imagine a population whose true mean is 50.

Take a sample of 20 observations.

```{r}

population <-

rnorm(

100000,

mean = 50,

sd = 10

)

sample_1 <-

sample(

population,

size = 20

)

mean(sample_1)

```

Now take another sample.

```{r}

sample_2 <-

sample(

population,

size = 20

)

mean(sample_2)

```

The means will not be identical.

This is sampling variability.

---

# 35. The Sampling Distribution

Repeat the sampling process many times.

```{r}

sample_means <-

replicate(

1000,

mean(

sample(

population,

size = 20

)

)

)

```

Plot the resulting means.

```{r}

tibble(

sample_mean = sample_means

) %>%

ggplot(

aes(sample_mean)

) +

geom_histogram(bins = 30) +

theme_minimal()

```

This is the **sampling distribution of the mean**.

---

# 36. Standard Error

The standard deviation of the sampling distribution is called the
**standard error**.

For a sample mean:

\[
SE =
\frac{\sigma}{\sqrt{n}}
\]

As sample size increases:

\[
SE \downarrow
\]

This is why larger samples generally provide more precise estimates.

---

# 37. The Central Limit Theorem

The **Central Limit Theorem** is one of the most important results in
statistics.

Under broad conditions, the sampling distribution of the mean approaches a
normal distribution as sample size increases.

This can happen even when the original population is not normally distributed.

---

# 38. Demonstrating the Central Limit Theorem

Create a strongly skewed population.

```{r}

skewed_population <-

rexp(

100000,

rate = 1

)

ggplot(

tibble(x = skewed_population),

aes(x)

) +

geom_histogram(bins = 50) +

theme_minimal()

```

The population is strongly right-skewed.

Now repeatedly sample from it.

```{r}

means <-

replicate(

1000,

mean(

sample(

skewed_population,

size = 30

)

)

)

```

Plot the sample means.

```{r}

tibble(

mean = means

) %>%

ggplot(

aes(mean)

) +

geom_histogram(bins = 30) +

theme_minimal()

```

The sampling distribution is much more approximately normal.

---

# 39. From Probability to Statistical Inference

We can now see the basic logic underlying statistical analysis.

We start with a probabilistic model.

For example:

```text
Population
    |
    | random sampling
    v
Sample
    |
    | calculate statistic
    v
Sample statistic
    |
    | probability theory
    v
Sampling distribution
    |
    | inference
    v
Conclusions about population
```

---

# 40. Parameters and Statistics

A **parameter** describes a population.

Examples:

\[
\mu = \text{population mean}
\]

\[
\sigma = \text{population standard deviation}
\]

A **statistic** describes a sample.

Examples:

\[
\bar{x} = \text{sample mean}
\]

\[
s = \text{sample standard deviation}
\]

We use sample statistics to learn about unknown population parameters.

---

# 41. Probability vs. Statistical Inference

There is an important distinction.

### Probability

We know something about the population and ask:

> What data are likely to occur?

### Statistical inference

We observe data and ask:

> What can these data tell us about the population?

This reversal is fundamental.

---

# 42. Likelihood

Suppose we observe some data.

A statistical model assigns probabilities to possible observations.

The **likelihood** asks:

> Given the observed data, how compatible are different parameter values with
> those data?

This idea underlies many statistical methods, including:

- linear regression
- logistic regression
- mixed-effects models
- generalized linear models

---

# 43. A Simple Example

Suppose we observe:

```text
7 successes out of 10 trials
```

We might ask:

> What value of p is most compatible with these data?

The likelihood is higher for values of \(p\) around 0.7 than for values close to
0.1.

---

# 44. Hypothesis Testing

Hypothesis testing asks whether our observed data would be unusual under a
specified null hypothesis.

For example:

\[
H_0: \mu_1 = \mu_2
\]

versus:

\[
H_1: \mu_1 \neq \mu_2
\]

The logic is probabilistic:

> If the null hypothesis were true, how surprising would our observed data be?

---

# 45. The p-value

A p-value is based on the probability of obtaining results at least as
extreme as those observed, under the null hypothesis.

It is **not**:

- the probability that the null hypothesis is true
- the probability that the results occurred by chance
- the probability that the alternative hypothesis is true

Understanding this distinction is essential.

---

# 46. Confidence Intervals

Probability theory also underlies confidence intervals.

A confidence interval describes uncertainty in an estimated parameter.

For example:

```text
Estimated mean = 52.4

95% CI = [49.8, 55.0]
```

The interval reflects sampling variability.

Larger samples generally produce narrower confidence intervals.

---

# 47. Statistical Models

A statistical model specifies assumptions about how observations are generated.

For example, a simple linear model can be written:

\[
Y_i = \beta_0 + \beta_1X_i + \epsilon_i
\]

where:

- \(Y_i\) is the observation
- \(X_i\) is a predictor
- \(\beta_0\) is the intercept
- \(\beta_1\) is the slope
- \(\epsilon_i\) represents unexplained variation

Mixed-effects models extend this framework by allowing parameters to vary
across participants and items.

---

# 48. Why This Matters for Psycholinguistics

Consider a word-rating experiment.

Suppose 50 participants rate 100 words on a 0–100 scale.

We observe:

```text
50 participants
×
100 words
=
5000 observations
```

But these observations are not 5000 completely independent pieces of
information.

They are generated by a structure involving:

- participants
- words
- experimental conditions
- lexical properties
- random variation

Probability theory provides the framework for reasoning about this uncertainty.

Mixed-effects models provide one way of representing that structure.

---

# 49. Exercises

## Exercise 1

Simulate 100 coin flips.

Calculate the proportion of heads.

Repeat this 1,000 times.

Plot the distribution of the resulting proportions.

---

## Exercise 2

Simulate samples of different sizes:

- 10
- 30
- 100
- 500

Compare the variability of the sample means.

---

## Exercise 3

Generate data from a highly skewed distribution.

Investigate how the distribution of sample means changes as sample size
increases.

---

## Exercise 4

A participant has a 75% probability of answering a question correctly.

Simulate 20 trials.

How many correct responses did the participant obtain?

Repeat this experiment 10,000 times and visualize the distribution.

---

## Exercise 5

Suppose the mean rating in a population is 60 and the standard deviation is
15.

Calculate the z-score for:

- 30
- 45
- 60
- 75
- 90

---

# 50. Summary

Probability theory provides the conceptual foundation for statistical
analysis.

The central ideas are:

- **Outcomes** are possible results of an experiment.
- **Events** are sets of outcomes.
- **Probability** describes uncertainty.
- **Conditional probability** describes probability given information.
- **Independence** describes situations in which one event does not affect
  another.
- **Random variables** represent uncertain quantities.
- **Probability distributions** describe how values are distributed.
- **Expected value** describes the long-run average.
- **Variance** describes variability.
- **Sampling distributions** describe how statistics vary across samples.
- **Standard errors** quantify sampling variability.
- **The Central Limit Theorem** explains why normal distributions appear so
  frequently in statistical inference.

The fundamental statistical logic is:

\[
\text{Population}
\rightarrow
\text{Sample}
\rightarrow
\text{Statistic}
\rightarrow
\text{Sampling Distribution}
\rightarrow
\text{Inference}
\]

Understanding this sequence makes later topics—including confidence
intervals, hypothesis tests, regression, and mixed-effects models—much easier
to understand.

---

# Further Topics

The natural next steps are:

1. Descriptive statistics
2. Sampling distributions
3. Confidence intervals
4. Hypothesis testing
5. t-tests
6. Correlation
7. Linear regression
8. Linear mixed-effects models
9. Generalized linear models
10. Bayesian statistics