---
title: "Visualizing Corpus Data"
author: "Introduction to R for Data Science"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

# Visualizing Corpus Data

## Introduction

In the previous tutorials you learned how to

- tokenize text
- remove stopwords
- count word frequencies
- calculate TF-IDF
- join lexical norms
- perform basic sentiment analyses

In this tutorial we'll focus on **visualization**.

Effective visualizations make it much easier to identify patterns in language
that would otherwise remain hidden.

By the end of this lesson you will be able to

- create publication-quality bar charts
- compare multiple corpora
- visualize lexical distributions
- create scatterplots
- build correlation heatmaps
- produce word clouds
- visualize word co-occurrence networks
- visualize document similarity using Principal Components Analysis (PCA)

---

# Packages

```{r message=FALSE}

library(tidyverse)
library(tidytext)
library(ggwordcloud)
library(widyr)
library(igraph)
library(ggraph)

```

---

# Example Corpus

```{r}

documents <- tibble(

doc = c(

"News",
"Sports",
"Literature"

),

text = c(

"The government introduced new economic reforms and discussed taxation.",

"The football team celebrated another victory after an exciting match.",

"The lonely traveler wandered quietly beneath the moonlight."

)

)

documents

```

---

# Tokenizing

```{r}

tokens <-

documents %>%

unnest_tokens(word, text)

tokens

```

---

# Removing Stopwords

```{r}

tokens_clean <-

tokens %>%

anti_join(stop_words,
          by = "word")

tokens_clean

```

---

# Word Frequencies

```{r}

word_counts <-

tokens_clean %>%

count(word,
      sort = TRUE)

word_counts

```

---

# Publication-quality Bar Charts

```{r}

word_counts %>%

slice_max(n,
          n = 10) %>%

ggplot(aes(reorder(word,
                   n),
           n)) +

geom_col() +

coord_flip() +

labs(

title = "Most Frequent Words",

x = "",

y = "Frequency"

) +

theme_minimal()

```

---

# Comparing Documents

```{r}

doc_counts <-

tokens_clean %>%

count(doc,
      word)

doc_counts

```

Plot the ten most frequent words in each document.

```{r}

doc_counts %>%

group_by(doc) %>%

slice_max(n,
          n = 10) %>%

ungroup() %>%

ggplot(

aes(reorder_within(word,
                   n,
                   doc),

    n)

) +

geom_col() +

facet_wrap(~doc,
           scales = "free") +

coord_flip() +

scale_x_reordered()

```

---

# Histograms

Suppose we have joined the Warriner norms.

```{r}

# Example only

warriner <-

tibble(

word = c("government",
         "victory",
         "lonely",
         "moonlight"),

valence = c(4.2,
            7.5,
            3.1,
            5.9),

arousal = c(5.3,
            7.0,
            4.6,
            3.8)

)

tokens_va <-

tokens_clean %>%

left_join(warriner,
          by = "word")

```

Visualize valence.

```{r}

ggplot(tokens_va,
       aes(valence)) +

geom_histogram(bins = 20) +

theme_minimal()

```

---

# Density Plots

Density plots provide a smoother view of distributions.

```{r}

ggplot(tokens_va,
       aes(valence)) +

geom_density() +

theme_minimal()

```

---

# Comparing Distributions

Compare valence across documents.

```{r}

ggplot(tokens_va,

aes(valence,
    fill = doc)

) +

geom_density(alpha = .4)

```

---

# Scatterplots

Compare two lexical variables.

```{r}

ggplot(tokens_va,

aes(valence,
    arousal)

) +

geom_point(size = 3)

```

Each point represents one word.

---

# Bubble Charts

Encode frequency with point size.

```{r}

bubble_data <-

tokens_va %>%

count(word,
      valence,
      arousal)

ggplot(

bubble_data,

aes(valence,
    arousal,
    size = n)

) +

geom_point(alpha = .7)

```

---

# Correlation Matrix

Suppose we have several lexical variables.

```{r}

lexical <-

tokens_va %>%

select(valence,
       arousal) %>%

drop_na()

cor(lexical)

```

Visualize correlations.

```{r}

corr <-

cor(lexical)

corr_df <-

as.data.frame(as.table(corr))

ggplot(corr_df,

aes(Var1,
    Var2,
    fill = Freq)

) +

geom_tile() +

geom_text(aes(label = round(Freq,2))) +

scale_fill_gradient2()

```

---

# Word Clouds

Word clouds emphasize frequent words.

```{r}

word_counts %>%

ggplot(

aes(

label = word,

size = n

)

) +

geom_text_wordcloud()

```

Word clouds are visually engaging, although they are less informative than bar
charts for precise comparisons.

---

# TF-IDF Visualization

```{r}

tfidf <-

tokens_clean %>%

count(doc,
      word) %>%

bind_tf_idf(word,
            doc,
            n)

```

```{r}

tfidf %>%

group_by(doc) %>%

slice_max(tf_idf,
          n = 8) %>%

ungroup() %>%

ggplot(

aes(

reorder_within(word,
               tf_idf,
               doc),

tf_idf

)

) +

geom_col() +

facet_wrap(~doc,
           scales = "free") +

coord_flip() +

scale_x_reordered()

```

---

# Word Co-occurrence Networks

Words appearing together often reveal important semantic relationships.

Create bigrams.

```{r}

bigrams <-

documents %>%

unnest_tokens(

bigram,

text,

token = "ngrams",

n = 2

)

```

Separate the two words.

```{r}

bigrams <-

bigrams %>%

separate(

bigram,

into = c("word1",
         "word2"),

sep = " "

)

```

Remove stopwords.

```{r}

bigrams <-

bigrams %>%

filter(

!word1 %in% stop_words$word,

!word2 %in% stop_words$word

)

```

Count word pairs.

```{r}

pairs <-

bigrams %>%

count(word1,
      word2,
      sort = TRUE)

pairs

```

---

# Network Graph

```{r}

graph <-

graph_from_data_frame(

pairs

)

ggraph(graph,
       layout = "fr") +

geom_edge_link() +

geom_node_point(size = 5) +

geom_node_text(aes(label = name),
               repel = TRUE)

```

This visualization reveals clusters of words that frequently occur together.

---

# Document Similarity

Suppose every row is a document.

Create a document-term matrix.

```{r}

dtm <-

tokens_clean %>%

count(doc,
      word) %>%

pivot_wider(

names_from = word,

values_from = n,

values_fill = 0

)

dtm

```

---

# Principal Components Analysis

Remove the document labels.

```{r}

matrix_data <-

dtm %>%

column_to_rownames("doc")

```

Run PCA.

```{r}

pca <-

prcomp(matrix_data,
       scale. = TRUE)

```

Plot.

```{r}

scores <-

as.data.frame(pca$x)

scores$doc <-

rownames(scores)

ggplot(

scores,

aes(PC1,
    PC2,
    label = doc)

) +

geom_point(size = 3) +

geom_text(vjust = -.5)

```

Documents that appear close together use similar vocabulary.

---

# Choosing the Right Visualization

| Goal | Recommended Plot |
|------|------------------|
| Word frequency | Bar chart |
| Lexical distribution | Histogram |
| Compare distributions | Density plot |
| Two lexical variables | Scatterplot |
| Frequency + lexical variables | Bubble chart |
| Correlations | Heatmap |
| Characteristic vocabulary | TF-IDF bar chart |
| Common phrases | Network graph |
| Document similarity | PCA |
| Exploratory overview | Word cloud |

---

# Best Practices

When creating visualizations:

- Always label your axes.
- Give every figure an informative title.
- Avoid unnecessary colors.
- Use `theme_minimal()` for clean figures.
- Order categories by frequency when possible.
- Prefer bar charts over word clouds when exact comparisons matter.

---

# Exercises

## Exercise 1

Create a bar chart showing the twenty most frequent words in your own corpus.

---

## Exercise 2

Join the Warriner norms and create histograms for

- Valence
- Arousal
- Dominance

How do the distributions differ?

---

## Exercise 3

Create a scatterplot of Valence versus Arousal.

Are emotionally positive words also highly arousing?

---

## Exercise 4

Calculate TF-IDF for your corpus and plot the ten highest-scoring words for each document.

---

## Exercise 5

Construct a word co-occurrence network using bigrams.

Which words form clusters?

---

## Exercise 6

If your corpus contains multiple documents, perform PCA and interpret the resulting clusters.

---

# Summary

In this tutorial you learned how to visualize corpus data using a variety of techniques.

You can now create:

- frequency bar charts
- histograms
- density plots
- scatterplots
- bubble charts
- correlation heatmaps
- TF-IDF visualizations
- word clouds
- co-occurrence networks
- PCA plots

Together with the previous tutorials, these techniques provide a solid foundation for exploratory corpus analysis and will prepare you for more advanced methods such as topic modelling, semantic embeddings, and statistical modelling.