---
title: "Introduction to R for Data Science"
output: html_document
---

# Introduction

R is a programming language designed for statistics, data analysis, and visualization.

## Installing Packages

```{r}
# install.packages("tidyverse")
```

## Loading Packages

```{r}
library(tidyverse)
```

## Set working directory
This code sets the directory to the current one.
```{r}
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)
```


# Basic Objects

```{r}
x <- c(1,2,3,4,5)
mean(x)
summary(x)
```

# Data Frames

```{r}
df <- tibble(
  name = c("Alice","Bob","Carol"),
  score = c(90,85,95)
)

df
```

## Save a data frame to a .csv
```{r}
write.csv(df,"../data/data.csv")
```


## Importing Data

```{r}
data <- read_csv("../data/data.csv")
head(data)
```

# Data Wrangling

```{r}
data %>%
  filter(score > 80) %>%
  arrange(desc(score))
```

# Visualization

```{r}
ggplot(df, aes(name, score)) +
  geom_col()
```

# Simple Statistics

```{r}
lm(score ~ 1, data = df)
```

# Exercises

1. Create a vector of 10 numbers.
2. Import a CSV file.
3. Filter rows using `filter()`.
4. Make a scatterplot with `ggplot2`.
5. Fit a linear model with `lm()`.
