---
title: "Analyzing a Similarity Rating Study"
author: "Your Name"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# load packages
library(tidyverse)
library(lme4)
library(lmerTest)   # for p-values
library(broom.mixed)

this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)
```

# Intro

a Similarity Rating study with a 4 × 1 design, where participants rate the similarity of word pairs.
Our primary goal is to predict similarity Rating from a predictor FeatureMatch using a linear mixed-effects regression model.

# Overview of statistical analysis

Statistical analysis often follows a common workflow:

    - Inspect the data (variables, distributions, missingness)

    - Visualize key relationships

    - Fit a statistical model that matches the design

    - Check model assumptions

    - Interpret coefficients and report results
    
    
# read in data

```{r}
# Replace with your actual data path or object
# Example: a CSV file with columns: Subject, WordPair, Rating, FeatureMatch, Condition

sim_data <- read_csv("data/similarity_ratings.csv")

# glimpse the structure
glimpse(sim_data)
```

```{r}
sim_data <- sim_data %>%
  mutate(
    # Ensure Rating is numeric
    Rating = as.numeric(Rating),
    # Make FeatureMatch a factor or numeric predictor as appropriate
    FeatureMatch = as.factor(FeatureMatch),
    WordPair = as.factor(WordPair),
    Subject = as.factor(Subject)    # if available
  )

summary(sim_data)
```


# Visualization

## histogram
```{r}

```


## barplot
```{r}
sim_data %>%
  ggplot(aes(x = FeatureMatch, y = Rating)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Similarity ratings by FeatureMatch level",
    x = "FeatureMatch",
    y = "Rating"
  )
```



```{r}
sim_data %>%
  group_by(WordPair, FeatureMatch) %>%
  summarize(
    mean_rating = mean(Rating, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = FeatureMatch, y = mean_rating)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ WordPair) +
  theme_minimal() +
  labs(
    title = "Mean similarity ratings by FeatureMatch and WordPair",
    x = "FeatureMatch",
    y = "Mean rating"
  )
```


# Model specification
```{r}
mod_wordpair <- lmer(
  Rating ~ FeatureMatch + (1 | WordPair),
  data = sim_data
)

summary(mod_wordpair)
```
## Report table

```{r}
report_table <- tidy_mod %>%
  select(term, estimate, std.error, conf.low, conf.high, p.value)

report_table
```



